---
title: "BT4BR Assignment 7"
author:
  name: Kaja Łucka
date: "2025-04-30" 
format: 
  html:
      df-print: paged                 # data frames as paginated tables
      tidy: false                     # disable automatic code tidying 
      embed-resources: true           # embed images in the HTML file
      theme: navbars                    # pretty theme :D
      toc: true                       # enable table of contents
      toc-location: right
      toc-depth: 6
      toc-expand: 6
      code-tools:
        source: true                  # show the source code
        toggle: true                  # allow the user to hide/show the code blocks
      code-fold: false                # by default code blocks are expanded
      code-summary: "Show the code"   # label the code folding toggle
      code-overflow: wrap             # wrap long lines of code
      page-layout: full               # use the full width of the page
editor: visual
editor_options: 
  markdown: 
    wrap: 72
execute:
  cache: true                         # code isn’t rerun unless it changes
---

*To complete this assignment I used materials shared with students
during "Basic Toolkit for Bioinformatics Research". I also intensely
browsed the internet and gathered information from the following
websites:*

[*dplyr.tidyverse.org/reference/summarise.html*](https://dplyr.tidyverse.org/reference/summarise.html)

[*dplyr.tidyverse.org/reference/filter.html*](https://dplyr.tidyverse.org/reference/filter.html)

[*quarto.org/docs/output-formats/html-themes.html*](https://quarto.org/docs/output-formats/html-themes.html)

## Task 1

Load the “iris” dataset and transform it to longer.

***TIP:** you should get 3 columns Species \| Trait \| Size*

```{r}
library(tidyr)
library(dplyr)
library(readr)

# loads iris dataset
data("iris")
iris

# transforms iris dataset into the longer format - takes columns which names contain "Sepal" or "Petal", puts the names into a new "Trait" column and the values into a new "Size" column
iris_long <- iris %>%
  pivot_longer(cols=matches("Sepal|Petal"), names_to="Trait", values_to="Size")
iris_long
```

## Task 2

Answer the following questions

***TIP:** you can answer each question “piping” only 2-3 commands*

### Task 2.1

What is the mean length and width of the sepal and petal of each
species?

```{r}
# groups iris_long by "Species" and "Trait" values, counts their mean and puts the result into the "Mean.Size" column
iris_long_means <- iris_long %>% 
  group_by(Species, Trait) %>% 
  summarise(Mean.Size = mean(Size), .groups = "keep")  # .groups = "keep" means that results stay grouped by both Species and Trait
iris_long_means
```

### Task 2.2

What is the largest and widest, petal and sepal, of virginica and
versicolor species?

```{r}
# takes data only for virginica and versicolor species from iris_long table, groups them by "Species" and "Trait" values, counts their maximum value and puts it into the "Maximum" column
iris_long_maximum <- iris_long %>% 
  filter(Species %in% c("virginica", "versicolor")) %>% 
  group_by(Species, Trait) %>% 
  summarise(Maximum = max(Size), .groups = "keep")  # .groups = "keep" means that results stay grouped by both Species and Trait
iris_long_maximum
```

### Task 2.3

For each species, for sepal and petal width and length, how many
observations are above the mean?

***TIP:** adding the means as column might help*

*Might seem difficult (it is not that difficult if you know the tricks
;) ). If you don’t manage to make it work, submit your best attempts and
we will discuss it in class.*

```{r}
# left joins iris_long with iris_long_means (every record from iris_long now has a corresponding mean value), leaves only records that are above mean (the value in the "Size" column is higher than the value in the "Mean.Size" column), then counts every row and saves the result in the "n" column
iris_long_above_mean <- iris_long %>% 
  left_join(y=iris_long_means, by = join_by(Species, Trait)) %>% 
  filter(Size > Mean.Size) %>% 
  group_by(Species, Trait) %>% 
  count()
iris_long_above_mean
```

## Task 3

Deliver a link to GitHub containing the a Quarto file and the compiled
HTML. It should contain explanations of what each chunk does. If you
have private repository, remember to send me invitation so I can access
it .
